name: dbi-exercises
class: polarnight

## DBI exercises

.pull-left[
- Load `mtcars` into the database from R
- Calculate the following in pgAdmin:
   * Average mileage by number of cylinders and type of transmission (hint: `?mtcars`)
   * Total weight of the listed cars per manufacturer (hint: look at rownames)
- Replicate the results using DBI in R
- Write an R function `query_avg` with
    * input: `con` (DBI connection), `name` (string), `n` (integer), and `m` (integer, optional)
    * output: SQL query for returning the average of column number `n` in table `name`, optionally grouped by column number `m`

    Hint: you need to use `dbListFields`
]

.pull-right[
- Verify that `query_avg` using the following code:
    ```{r ex1, eval = FALSE}
    q1 <- query_avg(con, "mtcars", 1)
    q2 <- query_avg(con, "mtcars", 4, 9)
    dbGetQuery(con, q1)
    dbGetQuery(con, q2)
    ```

    * Before running it: what do you expect it to return?
    * After running it: did it return the expectation?

- Delete the `mtcars` table from the database
- Ensure you only have the `nycflights13` tables left; check the output of
    ```{r ex2, eval = FALSE}
    dbListTables(con)
    ```

]

---

## Closing a connection

```{r dbi-close, error = TRUE}
dbDisconnect(con)
dbListTables(con)
```
