class: snowstorm, center, middle

# PostgreSQL in R

using the DBI package to query the database

---

## Setting up a connection

???
The basic component needed to use a database is a _connection object_, which contains all the information necessary for communicating with the postgres server

--

- Front end: `DBI` provides querying functions
- Back end: `RPostgreSQL` provides the postgres _driver_

???
You should already have these two packages installed.

--

```{r setup-connection}
library(DBI)
con <- dbConnect(
  drv      = RPostgreSQL::PostgreSQL(),  # driver
  host     = "localhost",                # host address (running locally on the computer)
  port     = 5432,                       # host port (default port)
  dbname   = "nycflights13",             # database name
  user     = "csor",                     # username for database owner
  password = "csor"                      # password for database owner
)
```

???
If your postgres server is set up with default settings, and you followed the guide in lecture 1 for setting up the `nycflights13` database, this is how to connect to it in R

---

## Inspect the database

```{r test-connection}
dbListTables(con)
str(dbGetInfo(con))
```

---
layout: true

## Querying

???
When we want to SELECT from the database, we have different options depending on when we want the results back.

---

```{r query-send-1}
(res <- dbSendQuery(con, "SELECT tailnum, year, manufacturer FROM planes LIMIT 5"))
(df <- dbFetch(res)); class(df)
dbClearResult(res)
```

???
## Query-Fetch-Clear

- `dbSendQuery`: submits and synchronously executes the SQL query to the database engine
  * NB: it does _not_ extract any records as an R object
  * The result is an object of class `PostgreSQLResult`, a so-called _result set_
- `dbFetch`: fetches the records from the result set
  * NB: this consumes the results
  * Repeating the fetch will result in no rows being returned the second time
- `dbClearResult`: frees up resources (locally and remotely) associated with a result set
  * Necessary to avoid exhausting memory etc.

---

```{r query-send-2}
res <- dbSendQuery(con, "SELECT tailnum, year, manufacturer FROM planes LIMIT 5")
while(!dbHasCompleted(res)) {
  chunk <- dbFetch(res, n = 2)
  print(nrow(chunk))
}
dbClearResult(res)
```

???
## Chunk-wise Fetch

`dbFetch` has an optional argument `n` that lets you control how many records you want to fetch.

These records are consumed from the result set, and the next time you fetch, you get records from where you left off.

Again, when using `dbSendQuery` and `dbFetch`, you need to clear the results afterwards manually.

---

```{r query-get}
(df <- dbGetQuery(con, "SELECT tailnum, year, manufacturer FROM planes LIMIT 5"))
class(df)
```

???
## All at once

The alternative to this 3-step process is to get everything at once with `dbGetQuery`.

Notice that you get _everything_ at once and not chunk-wise as with `dbFetch?`.

Most of the time, however, this is what you want.

---
layout: true

## Writing to the database

---

```{r write}
dbExistsTable(con, "iris")
dbWriteTable(con, "iris", iris)
dbExistsTable(con, "iris")
dbListFields(con, "iris")
```

???
`DBI` also lets us write to new or existing tables.

For this example:

- `dbExistsTable` tests for existence of a table
- `dbWriteTable` writes to a table
  * by default it creates a new table
  * extra arguments: `overwrite` and `append`

---

```{r append}
query <- 'SELECT "Species", count(*) FROM iris GROUP BY "Species"'
dbGetQuery(con, query)
dbWriteTable(con, "iris", iris, append = TRUE)
dbGetQuery(con, query)
```

???
Pay close attention to the quotes in the first line here;

- `Species` is quotes by double quotes
- The whole R string is quoted by single quotes
- In PostgreSQL, single quotes are for strings, and double quotes are for names
  * an unquoted name is case insensitive, in the way that it's converted to lower case
  * a quoted name is case sensitive
  * names with special symbols (spaces, punctuation, etc.) need to be quoted
- Everything went smoothly previously, because all names were lower case and had no special symbols in them

In a call to `dbWriteTable`, you can pass the argument `append = TRUE` to append rows to an existing table (would otherwise give an error).

Here, we just add the same rows again, causing the query to return double the counts from before.

---

```{r drop}
dbRemoveTable(con, "iris")
dbExistsTable(con, "iris")
dbListTables(con)
```

???
Deleting a table from the database is done with `dbRemoveTable`.

It returns `TRUE` on success and raises an error otherwise.

Be careful with this function `r emo::ji("wink")`
