class: center, middle, snowstorm

# tidyverse

a framework for doing data analysis with tidy data

???
Next, we turn to the `tidyverse` way of doing things.

This package exposes functions that, by now, should feel very SQL-like.
After a brief introduction to the core concepts of the package, we'll see that those functions translate _directly_ into SQL queries.

---

## Getting started

You should already have `tidyverse` installed; if not:

```{r tidyverse-install, eval = FALSE}
install.packages("tidyverse")
```

Load the most important packages in the collection with:

```{r tidyverse-load, message = TRUE}
library(tidyverse)
```

???
The `tidyverse` package is actually just a so-called _metapackage_.
It doesn't have any functionality of its own, but rather it acts as an umbrella over a set of packages each with an express purpose, all sharing an underlying design philosophy.

We'll only cover a very small subset of functionality in these slides, but an important one.

---

## Tibbles, the better data.frames

Run the two following lines:

```{r tibble, eval = FALSE}
iris
as_tibble(iris)
```

What's different?

???
Our good old `iris` dataset is just a regular `data.frame`, which, when printed, outputs as much as possible (up to 1000 rows).

Coercing it to a tibble gives an immediate quality-of-life improvement, in that the print method only shows what can be fit on the screen.

--

```{r tibble2}
class(iris)
class(as_tibble(iris))
```

???
We see that coercing to a tibble adds the classes `tbl` (pronounced "tibble") and `tbl_df`.

The `tbl` class is a broader, overarching class, and the `tbl_df` class is more specifically a "tibble dataframe" (which is usually what's referred to as a tibble).

---

## What is a tibble, actually?

--

.pull-left[
> Tibbles are data.frames that are lazy and surly: they do less and complain more
>
> [tibble.tidyverse.org](https://tibble.tidyverse.org/)
]

???
A tibble is a data.frame that expects you to think more about what you do;

- it is lazy, i.e. it doesn't change variable names or types, it doesn't do partial matching, and it doesn't drop dimensions
- it complains more, i.e. throws an error when a variable doesn't exit
- The idea behind this is that this forces you to confront problems early, leading to cleaner code
- NB: no row names, and row names are dropped on conversion (unless you specify they should be kept)

--

.pull-right[
```{r tibble-example}
tibble(
  x = 1:5,
  y = 1,
  z = x ^ 2 + y
)
```
]

???
Tibbles also have handy improvements when it comes to creation:

- it automatically recycles length 1 (and ONLY length 1) variables
- you can refer to variables you just created

---
layout: true

## Data analysis verbs

---
---

.right-column[
```{r dplyr-setup}
(mtcars_tbl <- as_tibble(rownames_to_column(mtcars, var = "model")))
```
]

---

.left-column[
### select
]

.right-column[
```{r dplyr-select}
select(mtcars_tbl, model, mpg, cyl, wt:am)
```

]
---

.left-column[
### select
### filter
]

.right-column[
```{r dplyr-filter}
filter(mtcars_tbl, am == 0, cyl != 8)
```

]

---

.left-column[
### select
### filter
### mutate
]

.right-column[
```{r dplyr-mutate}
tmp <- select(mtcars_tbl, model, mpg, wt)
mutate(tmp, new_colum = mpg / wt)
```

]

---

.left-column[
### select
### filter
### mutate
### arrange
]

.right-column[
```{r dplyr-arrange}
arrange(mtcars_tbl, desc(hp))
```

]

---

.left-column[
### select
### filter
### mutate
### arrange
### summarise
]

.right-column[
```{r dplyr-summarise}
summarise(mtcars_tbl, mean(mpg), max(hp), named_summary = first(cyl))
```

]

---

.left-column[
### select
### filter
### mutate
### arrange
### summarise
### group_by
]

.right-column[
```{r dplyr-groupby}
grouped_df <- group_by(mtcars_tbl, cyl)
summarise(grouped_df, avg_hp = mean(hp))
```
]

---
layout: false

## The pipe

`%>%`: uses the result of the left hand side as the first argument of the function on the right hand side.

In RStudio: `Ctrl`+`Shift`+`M` to insert `%>%`

--

.pull-left[
```{r pipe-before}
arrange(
  summarise(
    group_by(mtcars_tbl, cyl),
    avg_hp = mean(hp)
  ),
  desc(cyl)
)
```
]

--

.pull-right[
```{r pipe-after}
mtcars_tbl %>%
  group_by(cyl) %>%
  summarise(avg_hp = mean(hp)) %>%
  arrange(desc(cyl))
```
]

---

## Joins... with tibbles

.left-column[
- `inner_join`
- `left_join`
- `right_join`
- `full_join`
- `semi_join`
- `anti_join`
]

.right-column[
```{r dplyr-joins}
x <- tibble(key = 1:3, val = paste0("x", key))
y <- tibble(key = c(1, 2, 4), val = paste0("y", key))
```

.pull-left[
```{r dplyr-inner-join}
inner_join(x, y, by = "key")
semi_join(x, y, by = "key")
```

]

.pull-right[
```{r dplyr-full-join}
left_join(x, y, by = "key")
anti_join(x, y, by = "key")
```

]

]
