---
title: "Lecture 03 exercise solutions"
author: "Janus Valberg-Madsen"
output:
  html_notebook:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center")
```

## DBI exercises

```{r connection}
library(DBI)
con = dbConnect(RPostgreSQL::PostgreSQL(),
                dbname   = "nycflights13",
                user     = "csor",
                password = "csor")
```

```{r con-set, include = FALSE}
knitr::opts_chunk$set(connection = "con")
```


### 1 - Load `mtcars` into the database from R

```{r ex1}
dbWriteTable(con, "mtcars", mtcars)
```


### 2 - Calculate summaries with SQL

Average mileage by number of cylinders:

```{sql ex2-1}
SELECT cyl, am, avg(mpg) FROM mtcars
 GROUP BY cyl, am;
```

Total weight of the listed cars per manufacturer:

```{sql ex2-2}
SELECT left("row.names", strpos("row.names", ' ') - 1) AS manufacturer,
       sum(wt)
  FROM mtcars
 GROUP BY manufacturer;
```


### 3 - Replicate the results using R

Notice the choice of R string quotes and the escaped single quotes in `q2`:

```{r ex3}
q1 <- 'SELECT cyl, am, avg(mpg) FROM mtcars GROUP BY cyl, am;'
dbGetQuery(con, q1)
q2 <- '
SELECT left("row.names", strpos("row.names", \' \') - 1) AS manufacturer,
       sum(wt)
  FROM mtcars
 GROUP BY manufacturer;'
dbGetQuery(con, q2)
```


### 4 - Implementation of `query_avg`

```{r ex4}
query_avg <- function(con, name, n, m = NULL) {
  fields <- dbListFields(con, name)
  enquote <- function(x) paste0('"', x, '"')
  paste('SELECT',
        if (!is.null(m)) paste0(enquote(fields[m]), ','),
        'avg(', enquote(fields[n]), ') FROM', enquote(name),
        if (!is.null(m)) paste('GROUP BY', enquote(fields[m])), ';')
}
```


### 5 - Test of `query_avg`

Expectation:

1. `q1` should result in an error, not because of invalid syntax, but because the column number 1 is the automatically created `row.names` column - a string variable, for which an average cannot be calculated. You might expect `mpg` to be the first column, as it is for the `mtcars` data frame, but row names do not exist in SQL.
2. Keeping the above in mind, `q2` should return the average `disp` grouped by `vs`.

```{r ex5, error = TRUE}
q1 <- query_avg(con, "mtcars", 1)
dbGetQuery(con, q1)
q2 <- query_avg(con, "mtcars", 4, 9)
dbGetQuery(con, q2)
```


### 6 - Delete `mtcars` from the database again

```{r}
dbRemoveTable(con, "mtcars")
```

<!----------------------------------------------------------------------------->


## tidyverse exercises

```{r tidy-setup, message = FALSE}
library(tidyverse)
flights  <- tbl(con, "flights")
planes   <- tbl(con, "planes")
airlines <- tbl(con, "airlines")
airports <- tbl(con, "airports")
```


For each of the exercises, both pure SQL and tidyverse solutions are shown.


### 1 - Average arrival delay for each carrier

```{sql}
SELECT carrier, name, avg(arr_delay)
  FROM flights JOIN airlines USING (carrier)
 GROUP BY carrier, name;
```

```{r ex-tidy1}
q <- flights %>%
  inner_join(airlines, by = "carrier") %>%
  group_by(carrier, name) %>%
  summarise(mean(arr_delay, na.rm = TRUE))
q
show_query(q)
```

It is clear to see that while you can easily build valid SQL, it's not necessarily pretty SQL.


### 2 - Flights without records in `planes`

Using an _anti join_ subquery, we filter out the records that do have matches and count the remaining.

```{sql}
SELECT carrier, name, count(*)
  FROM flights JOIN airlines USING (carrier)
 WHERE NOT EXISTS (
   SELECT 1 FROM planes
    WHERE tailnum = flights.tailnum
 )
 GROUP BY carrier, name;
```

```{r ex-tidy2}
q <- flights %>%
  anti_join(planes, by = "tailnum") %>%
  inner_join(airlines, by = "carrier") %>%
  group_by(carrier, name) %>%
  count()
q
show_query(q)
```

Still, aside from the added verbosity, the two queries are largely the same.


### 3 - Number of flights by carrier and destination

```{sql}
SELECT airlines.name AS airline, airports.name AS airport, count(*)
  FROM flights
         JOIN airlines USING (carrier)
         JOIN airports ON (dest = faa)
 GROUP BY airline, airport
 ORDER BY airline, airport;
```

Sorted by airline and airport for clarity.

```{r ex-tidy3}
q <- flights %>%
  inner_join(airlines, by = "carrier") %>%
  inner_join(airports, by = c("dest" = "faa")) %>%
  rename(airline = name.x,
         airport = name.y) %>%
  group_by(airline, airport) %>%
  count() %>%
  arrange(airline, airport)
q
show_query(q)
```
